Efficient Data Provenance Analysis for Concurrent Programs

### 2014-03-20 paper.tex
checking in the edits I made to the introduction, related work, and background sections yesterday

\documentclass[preprint,9pt]{sigplanconf}
\usepackage{url}
\usepackage{listings}
\usepackage{subfigure}
\usepackage{times}
\usepackage{multirow}
\usepackage{color}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{color}
\usepackage{rotating}

@@

\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}\vspace{-2ex}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
\authorinfo{Anonymous for Submission}{}

@@

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to 
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically and each interleaving
can yield different behavior.  Reasoning about thread interactions is a central
challenge to building multi-threaded software because some thread interactions
have undesirable consequences, like crashing or degrading performance.

@@

A value's provenance is a description of the reason that it is stored in a
variable. Provenance information helps answer the question: "Why does this
variable contain this value?", which is central to debugging. Debugging a
concurrent program requires understanding how threads interact, and because  
communication happens via reads and writes to shared variables, {\em data
provenance information is essential}. Provenance can take on many forms, for
example, a value's provenance could be the instruction that wrote it, the
function that returned it, and in the context of concurrent programs, the
thread that produced it.

@@

Efficiently collecting debugging information, like data provenance, from production
systems is particularly important for concurrent software since bugs
manifest as failures rarely. When a production system fails and a programmer
receives a bug report, reproducing the reported failure to collect provenance
information may be difficult or impossible. It is essential to collect as much
information as performance margins permit from production, to reduce the need
for time-consuming failure reproduction.

Prior work has previously shown that some limited forms of provenance help
with debugging~\cite{badapples, whylineicse, tipslicingsurvey}. One effort
tracked code that wrote unusable values into variables~\cite{badapples}. That
work showed that revealing the connection between an unusable value's origin
and a failure it causes is helpful during debugging. Other work showed that
programmers debug more effectively when they can ask ``why'' questions about
values during debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea and are insufficient for dealing with concurrent program
bugs. First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}, which limits its applicability. 
% Limiting the applicability of their technique to only some failures, like {\tt
% Null Pointer Exceptions}.
Second, while prior work has shown that limited provenance tracking for Java
programs is possible with overheads low enough for production~\cite{badapples},
a more general form of provenance tracking for C and C++ programs has been
elusive. Prior approaches that build on slicing~\cite{thinslicing,
tipslicingsurvey, whylineicse} or heavy-weight memory checkers~\cite{badapples}
impose overheads too high for production for C and C++ programs. Third, no
prior technique that we know of has collected provenance information that is
tailored to debugging concurrent programs with overheads that are low enough
for use in production systems. Addressing concurrency is important because a
value could have originated in any of a large number of threads, making
reasoning about why a variable contains that value complex and difficult.

@@

In this work we address the challenges above and develop a novel form of
in-production provenance tracking for concurrent programs. We have three goals.
(1) To collect a general form of provenance information, not limited to only
certain values. (2) To provide provenance information that helps understand how
threads interact, which is the key to understanding concurrency bugs. (3) To
work with overheads low enough for use in production (ideally, less than 10\%),
especially for long-running server programs that experience a diversity of production behavior.

We develop a new provenance tracking mechanism called {\em last writer
slicing} that satisfies these goals.   As the program executes in production,
last writer slicing tracks the thread and program point that last wrote any
value to each potentially shared memory location --- its provenance.  Last
writer slices are saved with a core dump when the program fails and are
available in the debugger ({\em e.g.}, for breakpoint debugging).

@@

Motivated by these prior {\em ad hoc} efforts, we build a general mechanism
called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
(Section~\ref{sec:ctraps}). \ctraps can be used to implement several dynamic
analysis tailored to concurrent programs, like the ones discussed above.
\ctraps works by checking the last writer slice of each potentially shared
memory location when a thread accesses such a location. If the accessing
thread is different from the thread that last wrote the location, the access
signifies {\em communication} between the threads. When communication occurs,
\ctraps delivers a software {\em trap} to the accessing thread, giving it
information about how communication occurred. The trap can be handled by an
analysis-specific trap handler. Trap handlers implement {\em \ctraps
applications} through an extensible interface that can interpose on the
execution (Section~\ref{sec:apps}). By default, \ctraps has very low overheads
and similarly to last writer slicing are appropriate for use in production in
many cases. 
%\ctraps applications can perform arbitrary concurrency analyses,
%but at their own discretion and cost. 
As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.

@@

Data provenance information is directly useful during program debugging.  Prior
work~\cite{whylinechi,whylineicse} showed that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.  Origin tracking~\cite{badapples} and
program slicing~\cite{tipslicingsurvey} also make debugging easier by reporting
the provenance of values.  Like these prior efforts, in this work we focus on
using data provenance information for {\em debugging} -- understanding the root
cause of a failure -- as opposed to {\em bug detection} -- finding new, unknown
bugs.

@@

\paragraph{Debugging with Data Provenance Information}
Prior work has looked at various forms of data provenance for debugging.

We have discussed bad value origin tracking~\cite{badapples} throughout this
work and compared directly with that technique in our evaluation.  Bad value
origin tracking is like our work because they store provenance information to
be used for debugging -- a similar goal to ours.  Bad value origin tracking
differs from our work in several ways.  First, their technique only tracks the
provenance of {\em unusable} values -- a fraction of all values in an
execution.  Our work, instead, tracks provenance for any value stored in a
potentially shared memory location, which includes a broader set of provenance
and targets our technique to concurrency debugging.  Second, their technique
does not run with overheads tolerable for performance for C and C++ programs.
They implement their technique for C and C++ programs using Valgrind MemCheck,
which has high runtime overheads.  Our technique has overheads low enough for
production in many cases, especially in server applications.  Third, their
technique does not explicity record thread information with tracked origins.
Our work focuses on concurrency and we include thread information.  As we show
in Section~\ref{sec:eval:debugging} thread information is important for
understanding some bugs.  Fourth, their technique uses bad value piggy-backing
to store origins, we use a shadow memory.

Like our work, the WhyLine debugger~\cite{whylineicse, whylinechi} allows
programmers to ask questions about why variables hold particular values during
debugging.  WhyLine answers by presenting data- and control-flows that
influenced those variables.  WhyLine differs from our work in that it uses
dependence slicing on recorded executions to answer provenance questions and
has overheads too high for production use.  WhyLine also did not explicitly
focus on concurrency.

@@

\paragraph{Communication Tracking in Concurrent Programs}
Several techniques have been proposed do some form of dependence and
communication tracking to analyze concurrent programs.

>>>>

\documentclass[preprint,10pt]{sigplanconf}
\usepackage{url}
\usepackage{listings}
\usepackage{subfigure}
\usepackage{times}
\usepackage{multirow}
\usepackage{color}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{color}
\usepackage{rotating}

@@

\newcommand{\Caption}[1]{\begin{minipage}{1.0\columnwidth} \caption{#1} \end{minipage}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
\authorinfo{Anonymous for Submission}{}

@@

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to sequential
reasoning, multi-threaded programs require complex reasoning about many threads
of control and their interactions.  In shared memory programs, threads interact
by reading and writing memory locations.  Threads' reads and writes interleave
arbitrarily and nondeterministically. Each interleaving can lead to different
behaviors and memory states, some of which may be undesirable, like crashes or
data corruption.  Moreover, because complete concurrency testing is infeasiable
in general, undesirable interleavings can evade testing and lead to production
failures.

The problem we are addressing in this work is that debugging support for
understanding thread interactions is inadequate.  Today, programmers use a
debugger to examine memory contents either during a debugging execution or by
loading a core dump from a failed execution.  Examining the contents of memory
reveals {\em what} the state of the program is ({\em e.g.}, the values of
variables).  Unfortunately, the real question programmers must answer is {\em
why} the state of the program is what it is.  Understanding why a program
entered an undesirable state ({\em e.g.}, crashed) is the key to changing the
code to prevent future executions from entering that state.  Multi-threading
makes this understanding elusive because implicating particular operations in
particular threads for a failure is a complex task.  Aiding developers in this
understanding for failed production executions is important because
hard-to-find bugs may manifest in production only.

@@

Our approach to solving that problem is to provide the programmer with {\em
provenance} information that can be examined alongside the program's memory
state ({\em i.e.}, core dump).  A value's provenance describes why a memory
location contains its value.  In this work we develop a new provenance tracking
mechanism called {\em last writer slicing} that satisfies these goals.  A
memory location's last writer slice is a record containing the program point
and the identifier of the thread that last wrote data to that memory location.
As the program executes in production, we collect last writer slices for each
potentially shared memory location and those last writer slices record the
provenance of the values stored in those locations.  Last writer slices are
saved with a core dump when the program fails and are available in the debugger
({\em e.g.}, for breakpoint debugging).

We have three goals for last writer slicing:  (1) To collect a general form of
provenance information, not limited to certain values or locations only. (2) To
provide provenance information focused on how threads interact to aid in
understanding concurrency bugs. (3) To engineer our system to have overheads
low enough for use in production, aiming for a 10 -- 50\% maximum overhead.

@@

@@

@@

Motivated by these prior {\em ad hoc} efforts, we build a general mechanism
called {\em Communication Traps} or {\em \ctraps} on top of last writer slices
(Section~\ref{sec:ctraps}). \ctraps is useful for implementing dynamic
concurrency analyses, like the ones discussed above.  \ctraps runs alongside a
program and uses last writer slices to identify memory accesses that lead to
{\em communication} between the threads. Communcation occurs when a thread
accesses memory last written by a different thread and \ctraps delivers a {\em
trap} to a thread when it communicates with another thread, giving it
information about how communication occurred.  Programmer-defined {\em trap
handlers} can implement arbitrary analysis using that information. By default,
\ctraps has low overheads that are appropriate production in many cases.  As we
show in our evaluation, the overhead of \ctraps applications scales with the
complexity of their analysis.

@@

Prior work shows that some forms of data provenance information is directly
useful during program debugging.  The WhyLine
Debugger~\cite{whylinechi,whylineicse} showed that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.  Bad Value Origin
Tracking~\cite{badapples} is a limited form of provenance analysis that tracks
intructions that stored certain unusable values.  The authors of Bad Value Origin Tracking showed that even such limited provenance information
helps understand some bug root causes. Like these
prior efforts, we focus on using provenance analysis for
{\em debugging} -- understanding the root cause of a failure -- as opposed to
{\em bug detection} -- finding new, unknown bugs.

@@

\paragraph{Debugging with Data Provenance Information}
Prior work has shown that some limited forms of provenance help
with debugging~\cite{badapples, whylineicse, tipslicingsurvey}.

Bad Value Origin Tracking~\cite{badapples} tracked code that wrote unusable
values into variables~\cite{badapples}. That work showed that revealing the
connection between an unusable value's origin and a failure it causes is
helpful during debugging.  Bad value origin tracking is the closest prior work
to this work~\cite{badapples} and we directly compare to it in our evaluation.
At a high level, bad value origin tracking is like our work because they track
provenance information for debugging, which is a similar goal to ours.

Bad value origin tracking relates to our work in several other ways.  First,
their technique tracks the provenance of {\em unusable} values only -- a
fraction of all values in an execution.  Our work, instead, tracks provenance
for any value stored in a potentially shared memory location, which includes a
broader set of provenance and targets our technique to concurrency debugging.
Second, for Java programs, bad value origin tracking has runtime overheads
acceptable for production, which was a goal for our system.  Bad value origin
tracking ``piggybacks'' provenance information on storage allocated by the Java
runtime for bad values.  Piggybacking is a key to their low overheads.  For
C/C++, the overhead of bad value origin tracking is much higher.  In C/C++,
they cannot use piggybacking and, instead, use heavy-weight binary
instrumentation with overheads unacceptable for production.  Our technique has
overheads low enough for production for C/C++ programs, especially for server
applications.  Third, their technique does not explicity record thread
information with tracked origins.  Our work focuses on concurrency and we
include thread information.  As we show in Section~\ref{sec:eval:debugging}
thread information is important for some bugs.

Other work showed that programmers debug more effectively when they can ask
``why'' questions about values during
debugging~\cite{tipslicingsurvey,whylineicse}.  Like our work, the WhyLine
debugger~\cite{whylineicse, whylinechi} allows programmers to ask questions
about why variables hold particular values during debugging.  WhyLine answers
by presenting data- and control-flows that influenced those variables.  WhyLine
differs from our work in that it uses slicing on recorded executions to answer
provenance questions, it has high run time overheads, and does not focus on
concurrency.

@@

\paragraph{Communication and Sharing Analyses in Concurrent Programs}
Several techniques have been proposed analyze dependences and
communication in concurrent programs.

A closely related recent effort in this space is Octet~\cite{octet} that
captures inter-thread communication soundly and, like \ctraps, can be used to
build dynamic concurrency analyses.  Octet differs from \ctraps in that it is
sound and was built to support analyses on Java programs.  Additionally, Octet
tracks the ``locality state'' of memory locations, rather than the last writer
slice information.  Octet's average run time overhead (26\%) is comparable to
the overhead of both last writer slicing and \ctraps and all three techniques
have low enough overheads in many cases for production use.
